**OptiFit: a fast method for fitting amplicon sequences to existing OTUs**

### 2021-08-27 paper/manuscript.tex
adjustments from pat comments

Colorectal cancer is one of the most common cancers in men and women and
a leading cause of cancer related deaths in the United States
(\protect\hyperlink{ref-siegel2020}{1}). Early detection and treatment
are essential to increase survival rates, but for a variety of reasons
including the invasiveness and high cost of screening
(i.e.~colonoscopy), many people do not comply with recommended screening
guidelines (\protect\hyperlink{ref-garcuxeda2011a}{2}) prompting a need
for low cost, non-invasive detection methods. A growing body of research
points to the gut microbiome as a promising target for non-invasive
screening to detect screen relevant neoplasias (neoplasms?) (SRNs)
consisting of pre-cancerous polyps (i.e.~advanced adenomas) and
carcinomas. Efforts to realize the diagnostic potential of the gut
microbiome in detecting SRNs have focused on machine learning (ML)
methods using abundances from operational taxonomic unit (OTU)
classifications based on amplicon sequencing of the 16S rRNA gene \{\}.
However, whether this is the optimal taxonomic resolution for
classifying SRNs from microbiome data is unknown. Additionally, recent
work has pushed for the use of amplicon sequence variants (ASVs) to
replace OTUs for marker-gene analysis because of the improved resolution
with ASVs (\protect\hyperlink{ref-callahan2017}{3}). However, whether
the additional resolution provided by ASVs is useful for ML
classification is unclear \{\}. Since ML classification relies on
consistent differences between groups, its possible that the resolution
at the ASV level is too individualized to accurately differentiate
groups. Topçuoğlu \emph{et al}
(\protect\hyperlink{ref-topuxe7uolu2020}{4}) recently demonstrated
effective application of machine learning (ML) to microbiome based
classification problems and developed a framework for applying ML
practices in a more reproducible way (mikropml). This analysis utilizes
the reproducible framework developed by Topçuoğlu \emph{et al} to
quantify which ML method and taxonomic level produce the best performing
classifier for detecting SRNs from microbiome data.

Utilizing publicly available 16S rRNA sequence data from stool of
patients with SRNs and healthy controls, we generated abundance tables
with Mothur \{\} annotated to Phylum, Class, Order, Family, Genus, OTU
and ASV levels. Using the taxonomic abundance data, we quantified how
accurately samples could be classified as ``normal'' or ``SRN''
(i.e.~advanced adenoma or carcinoma) using five machine learning methods
with the Mikropml R package (methods). Across the five machine learning
methods tested, model performance tended to increase with taxonomic
level usually peaking around genus/OTU level before dropping off
slightly with ASVs (Supplemental Figure 1). Random forest (RF) was
consistently the top performer at most taxonomic levels. Within the RF
model, the highest AUROCs were observed for family (median AUROC:
0.687), genus (median AUROC: 0.686), and OTU (median AUROC: 0.698) level
data with no significant difference between the three (Figure 1A,
Supplemental Figure 2). Performance with ASVs (median AUROC: 0.676) was
significantly lower than OTUs (p \textless{} 0.01), but borderline
equivalent to family (p = 0.06) and genus (p = 0.05) levels (Figure 1A).
These results were consistent for ASVs generated with DADA2
(\protect\hyperlink{ref-callahan2016}{5}) (median AUROC: 0.66). These
results suggest that finer taxonomic resolution is not necessarily
better for identifying individuals with SRNs based on microbiome
composition.

While comparing AUROC values between models is a useful way to assess
the overall model performance, they summarize the performance across all
thresholds and can be misleading since models with the same AUC can have
different ROC curve shapes (\protect\hyperlink{ref-lobo2008}{6}).
Depending on the intended implementation of the model, one may want to
optimize the true-positive rate (or sensitivity) over the false-positive
rate (or 1-specificity), or vice versa. In this case, the optimal model
will detect as many true positives (people with SRNs) as possible. To
further compare the model performance across taxonomic levels we
compared the sensitivity of the models at a specificity of 90\%. The
highest sensitivity values were observed for Family, Genus, and OTU
level data (Figure 1B), consistent with the AUROC results. Phylum,
Class, Order, and ASV sensitivity values were all significantly lower
than Family, Genus, and OTU sensitivity values (Figure 1B). This
analysis further supports that finer resolution is does not improve SRN
detection.

One hypothesis for the observation that model performance increases from
phylum to OTU level then drops slightly at ASV level is that at higher
taxonomic levels (e.g.~phylum) there are too few taxa and too much
overlap to reliably differentiate between cases and controls. At the
level of genus/OTU data there is enough data and variation but at the
ASV level, the data is too specific to individuals and doesn't overlap
enough. Examination of the prevalence of taxa in samples at each level
supports this idea. A majority of taxa are present in greater than 75\%
of samples at the phylum (67\% of taxa) and class (63\% of taxa) levels.
The opposite is observed at the OTU and ASV level where 60\% and 53\% of
taxa respectively are only present in less than 25\% of samples
(Supplemental Figure 3). Of note, the ML pipeline includes a
pre-processing step that occurs prior to training and classifying the ML
models (methods). As part of this step, features are removed that will
not provide useful information to build the model. For example,
perfectly correlated features provide the same information to build the
model and thus can be collapsed. Additionally, features with zero or
near-zero variance will not help the model differentiate groups and thus
can be removed. Interestingly, despite starting with 104106 features at
the ASV level, only 478 (0.5\%) remained after pre-processing. At the
OTU level, 20079 of the 705 features (3.5\%) remained after
preprocessing (Table 1). While the resolution provided by ASVs is useful
in certain contexts \{\}, these results suggest that the resolution is
too fine for use in machine learning classification of SRNs based on
microbiome composition.

A look into the most important taxa at each level for classifying
samples reveals some nesting where several genera and their higher
taxonomic classifications are in the top 10 most important taxa
(Supplemental Figure 4). For example, the genus \emph{Gemella} is an
important taxa at the genus and OTU levels and its higher
classifications are also important (\emph{Firmicutes} \textgreater{}
\emph{Bacilli} \textgreater{} \emph{Bacillales} \textgreater{}
\emph{Bacillales Incertae Sedis XI} \textgreater{} \emph{Gemella}).
\emph{Fusobacterium} displays a similar pattern, except that the family
level classification (\emph{Fusobacteriaceae}) importance is ranked 16th
out of 54 families, a little beyond the top 10. In the case of
unclassified \emph{Lachnospiraceae}, there are several OTUs with this
label that are in the top 10, however at the Genus level this taxon is
ranked lower in importance (21st out of 115 genera) suggesting there may
be some benefit to separating different taxonomic groupings within
\emph{Lachnospiraceae}.

These results demonstrate that consideration of the appropriate
taxonomic resolution for utilizing the microbiome as a predictive tool
is warranted. In general, we found that finer taxonomic resolution
(e.g.~OTU and ASV) does not add additional sensitivity to predicting
SRNs based on microbiome composition. Additionally, at the ASV level the
fine resolution actually impedes model performance due to the sparsity
of shared taxa and leads to decreased model performance. The tendancy
for ASV level annotation to split single bacterial genomes into multiple
taxa \{Pat mSphere 2021\} could also be a contributing factor to the
sparsity of shared taxa. Overall, either Family, Genus, or OTU level
taxonomy appear to perform equally for predicting subjects with SRNs
based on the composition of the gut microbiome. A potential benefit of
utilizing genus or family level data could be that it may allow for
merging data generated from different 16S regions or sequencing
platforms.

@@

\textbf{\emph{Data processing.}} Sequence data was processed with Mothur
(1.44.3) using the SILVA reference database (v132) to produce count
tables for phylum, class, order, family, genus, OTU, and ASV following
the Schloss Lab MiSeq SOP described on the Mothur website \{\}. ASV
level data was also produced using DADA2 to ensure consistent results
with a different pipeline. Data was processed following the DADA2
pipeline, but setting pool=TRUE to infer ASVs from the whole dataset
rather than per sample. The resulting ASV table was subsampled for
consistency with the Mothur data.

\textbf{\emph{Machine Learning.}} Machine learning models were run with
the R package mikropml (v0.0.2) \{\} to predict the diagnosis category
(normal vs SRN) of each sample. Data was preprocessed to normalize
values (scale/center), remove values with zero or near-zero variance,
and collapse collinear features using default parameters. Initially the
models were run with default hyperparameters, but were expanded if the
peak performance was at the edge of the hyperparameter range. Each
taxonomic model taxonomic level combination (e.g.~Random Forest on
genus) was run with 100 different seeds (1-100). Each seed split the
data into a training (80\%) and testing (20\%) set, and output
performance of the training and testing as area under the receiver
operating curve (AUROC).

@@

\includegraphics{manuscript_files/figure-latex/fig1-1.pdf}

@@

\begin{longtable}[]{@{}lrrc@{}}
\toprule
Taxonomic Level & Number of Features &
\makecell[c]{Number of Features \\ After Preprocessing} &
\makecell[c]{Percent of Features Kept \\ After Preprocessing} \\ \addlinespace
\midrule
\endhead
Phylum & 19 & 9 & 47.4 \% \\ \addlinespace
Class & 36 & 19 & 52.8 \% \\ \addlinespace
Order & 65 & 28 & 43.1 \% \\ \addlinespace
Family & 124 & 54 & 43.5 \% \\ \addlinespace
Genus & 316 & 115 & 36.4 \% \\ \addlinespace
OTU & 20079 & 705 & 3.5 \% \\ \addlinespace
ASV & 104106 & 478 & 0.5 \% \\ \addlinespace
\bottomrule
\end{longtable}

@@

\includegraphics{../exploratory/figures/all_model_level.png}

@@

\includegraphics[width=\textwidth,height=0.5\textheight]{../exploratory/figures/roc_by_level.png}

@@

\includegraphics[width=\textwidth,height=0.45\textheight]{../exploratory/figures/bin_prevalence.png}\\
\textbf{Supplemental Figure 3: Prevalence of Taxa in Samples.}
Distribution of taxa prevalence in samples at each taxonomic level.

@@

\includegraphics[width=\textwidth,height=0.85\textheight]{../exploratory/figures/top10_imp.png}\\
\textbf{Supplemental Figure 4: Top 10 important taxa at each taxonomic
level.} Summary of the 10 most important taxa for the random forest
models at each taxonomic level based on the average decrease in AUC when
the feature is permuted.

@@

\leavevmode\hypertarget{ref-callahan2016}{}%
\CSLLeftMargin{5. }
\CSLRightInline{\textbf{Callahan BJ}, \textbf{McMurdie PJ},
\textbf{Rosen MJ}, \textbf{Han AW}, \textbf{Johnson AJA}, \textbf{Holmes
SP}. 2016. DADA2: High-resolution sample inference from Illumina
amplicon data. Nature Methods \textbf{13}:581--583.
doi:\href{https://doi.org/10.1038/nmeth.3869}{10.1038/nmeth.3869}.}

\leavevmode\hypertarget{ref-lobo2008}{}%
\CSLLeftMargin{6. }
\CSLRightInline{\textbf{Lobo JM}, \textbf{Jiménez-Valverde A},
\textbf{Real R}. 2008. AUC: a misleading measure of the performance of
predictive distribution models. Global Ecology and Biogeography
\textbf{17}:145--151.
doi:\href{https://doi.org/10.1111/j.1466-8238.2007.00358.x}{10.1111/j.1466-8238.2007.00358.x}.}

>>>>

Colorectal cancer is one of the most common cancers in men and women and
a leading cause of cancer related deaths in the United States
(\protect\hyperlink{ref-siegel2020}{1}). Early detection and treatment
are essential to increase survival rates, but for a variety of reasons
including the invasiveness and high cost of screening
(i.e.~colonoscopy), many people do not comply with recommended screening
guidelines (\protect\hyperlink{ref-garcuxeda2011a}{2}) prompting a need
for low cost, non-invasive detection methods. A growing body of research
points to the gut microbiome as a promising target for non-invasive
screening to detect screen relevant neoplasia (SRNs) consisting of
pre-cancerous polyps (i.e.~advanced adenomas) and carcinomas. Efforts to
realize the diagnostic potential of the gut microbiome in detecting SRNs
have focused on machine learning (ML) methods using abundances of
operational taxonomic unit (OTU) classifications based on amplicon
sequencing of the 16S rRNA gene \{\}. However, whether this is the
optimal taxonomic resolution for classifying SRNs from microbiome data
is unknown. Additionally, recent work has pushed for the use of amplicon
sequence variants (ASVs) to replace OTUs for marker-gene analysis
because of the improved resolution with ASVs
(\protect\hyperlink{ref-callahan2017}{3}). However, whether the
additional resolution provided by ASVs is useful for ML classification
is unclear \{\}. Since ML classification relies on consistent
differences between groups, its possible that the resolution at the ASV
level is too individualized to accurately differentiate groups.
Topçuoğlu \emph{et al} (\protect\hyperlink{ref-topuxe7uolu2020}{4})
recently demonstrated how to effectively apply machine learning (ML)
methods to microbiome based classification problems and developed a
framework for applying ML practices in a more reproducible way
(mikropml). This analysis utilizes the reproducible framework developed
by Topçuoğlu \emph{et al} to determine which ML method and taxonomic
level produce the best performing classifier for detecting SRNs from
microbiome data.

Utilizing publicly available 16S rRNA sequence data from stool of
patients with SRNs and healthy controls, we generated abundance tables
with mothur \{\} annotated to phylum, class, order, family, genus, OTU
and ASV levels. Using the taxonomic abundance data, we quantified how
reliably samples could be classified as ``normal'' or ``SRN''
(i.e.~advanced adenoma or carcinoma) using five machine learning methods
with the mikropml R package (methods). Across the five machine learning
methods tested, model performance tended to increase with taxonomic
level usually peaking around genus/OTU level before dropping off
slightly with ASVs (Supplemental Figure 1). Regardless of the taxonomic
level, random forest (RF) models consistently had the largest area under
the receiver operating characteristic curve (AUROC). Within the RF
model, the highest AUROCs were observed for family (median AUROC:
0.687), genus (median AUROC: 0.686), and OTU (median AUROC: 0.698) level
data with no significant difference between the three (Figure 1A,
Supplemental Figure 2). Performance with ASVs (median AUROC: 0.676) was
significantly lower than OTUs (p \textless{} 0.01), but comparable to
family (p = 0.06) and genus (p = 0.05) levels (Figure 1A). These results
suggest that finer taxonomic resolution is not necessarily better for
identifying individuals with SRNs based on microbiome composition.

While comparing AUROC values between models is a useful way to assess
the overall model performance, they summarize the performance across all
thresholds and can be misleading since models with the same AUC can have
different ROC curve shapes (\protect\hyperlink{ref-lobo2008}{5}).
Depending on the intended implementation of the model, one may want to
optimize the true-positive rate (or sensitivity) over the false-positive
rate (or 1-specificity), or vice versa. In this case, the optimal model
will detect as many true positives (people with SRNs) as possible. To
further compare the model performance across taxonomic levels we
compared the sensitivity of the models at a specificity of 90\%. The
highest sensitivity values were observed for family, genus, and OTU
level data (Figure 1B), consistent with the AUROC results. phylum,
class, order, and ASV sensitivity values were all significantly lower
than family, genus, and OTU sensitivity values (Figure 1B). This
analysis further supports the observation that finer resolution is does
not improve SRN detection.

One hypothesis for the observation that model performance increases from
phylum to OTU level then drops slightly at ASV level is that at higher
taxonomic levels (e.g.~phylum) there are too few taxa and too much
overlap to reliably differentiate between cases and controls. At the
level of genus or OTU there is enough data and variation but at the ASV
level, the data is too specific to individuals and does not overlap
enough. Examination of the prevalence of taxa in samples at each level
supports this idea. A majority of taxa were present in greater than 75\%
of samples at the phylum (67\% of taxa) and class (63\% of taxa) levels.
The opposite was observed at the OTU and ASV level where 60\% and 53\%
of taxa respectively were only present in less than 25\% of samples
(Supplemental Figure 3). Of note, the ML pipeline includes a
pre-processing step that occured prior to training and classifying the
ML models. As part of this step, taxa were removed that would not
provide useful information to build the model. For example, perfectly
correlated taxa provide the same information to build the model and thus
can be collapsed. Additionally, features with zero or near-zero variance
across samples would not help the model differentiate groups and thus
can be removed. Interestingly, despite starting with 104,106 ASVs, only
478 (0.5\%) remained after pre-processing. At the OTU level, 705 of the
20,079 OTUs (3.5\%) remained after preprocessing (Table 1). While the
resolution provided by ASVs is useful in certain contexts \{\}, these
results suggest that the resolution is too fine for use in machine
learning classification of SRNs based on microbiome composition.

A look into the most important taxa at each level for classifying
samples revealed some nesting where several genera and their higher
taxonomic classifications were in the top 10 most important taxa
(Supplemental Figure 4). For example, the genus \emph{Gemella} was an
important taxon at the genus and OTU levels and its higher
classifications were also important (\emph{Firmicutes} \textgreater{}
\emph{Bacilli} \textgreater{} \emph{Bacillales} \textgreater{}
\emph{Bacillales Incertae Sedis XI} \textgreater{} \emph{Gemella}).
\emph{Fusobacterium} displayed a similar pattern, except that the family
level classification (\emph{Fusobacteriaceae}) importance was ranked
16th out of 54 families. In the case of unclassified
\emph{Lachnospiraceae}, there were several OTUs with this label that
were in the top 10, however at the genus level this taxon was ranked
lower in importance (21st out of 115 genera) suggesting there may be
some benefit to separating different taxonomic groupings within
\emph{Lachnospiraceae}.

These results demonstrate that consideration of the appropriate
taxonomic resolution for utilizing the microbiome as a predictive tool
is warranted. In general, we found that finer taxonomic resolution
(e.g.~OTU and ASV) did not add additional sensitivity to predicting SRNs
based on microbiome composition. Additionally, at the ASV level the fine
resolution actually impedes model performance due to the sparsity of
shared taxa and led to decreased model performance. The tendancy for ASV
level annotation to split single bacterial genomes into multiple taxa
\{Pat mSphere 2021\} could also be a contributing factor to the sparsity
of shared taxa. Overall, either family, genus, or OTU level taxonomy
appear to perform equally for predicting subjects with SRNs based on the
composition of the gut microbiome. A potential benefit of utilizing
genus or family level data could be that it may allow for merging data
generated from different 16S rRNA gene regions or sequencing platforms.

@@

\textbf{\emph{Data processing.}} Sequence data was processed with mothur
(1.44.3) using the SILVA reference database (v132) to produce count
tables for phylum, class, order, family, genus, OTU, and ASV following
the Schloss Lab MiSeq SOP described on the mothur website. ASV level
data was also produced using DADA2 to ensure consistent results with a
different pipeline. Data was processed following the DADA2 pipeline, but
setting pool=TRUE to infer ASVs from the whole dataset rather than per
sample. The resulting ASV table was subsampled for consistency with the
mothur data. These results were consistent for ASVs generated with DADA2
(\protect\hyperlink{ref-callahan2016}{6}) (median AUROC: 0.66).

\textbf{\emph{Machine Learning.}} Machine learning models were run with
the R package mikropml (v0.0.2) \{\} to predict the diagnosis category
(normal vs SRN) of each sample. Data was preprocessed to normalize
values (scale/center), remove values with zero or near-zero variance,
and collapse collinear features using default parameters. Initially the
models were run with default hyperparameters, but were expanded if the
peak performance was at the edge of the hyperparameter range. Each
taxonomic model taxonomic level combination (e.g.~random forest on
genus) was run with 100 different seeds. Each seed split the data into a
training (80\%) and testing (20\%) set, and output performance of the
training and testing as area under the receiver operating curve (AUROC).

@@

\includegraphics{figure_1.png}

@@

\begin{longtable}[]{@{}lrrc@{}}
\toprule
Taxonomic Level & Number of Features &
\makecell[c]{Number of Features \\ After Preprocessing} &
\makecell[c]{Percent of Features Kept \\ After Preprocessing} \\
\midrule
\endhead
Phylum & 19 & 9 & 47.4 \% \\
Class & 36 & 19 & 52.8 \% \\
Order & 65 & 28 & 43.1 \% \\
Family & 124 & 54 & 43.5 \% \\
Genus & 316 & 115 & 36.4 \% \\
OTU & 20,079 & 705 & 3.5 \% \\
ASV & 104,106 & 478 & 0.5 \% \\
\bottomrule
\end{longtable}

@@

\includegraphics{figure_s1.png}

@@

\includegraphics[width=\textwidth,height=0.5\textheight]{figure_s2.png}

@@

\includegraphics[width=\textwidth,height=0.45\textheight]{figure_s3.png}\\
\textbf{Supplemental Figure 3: Prevalence of Taxa in Samples.}
Distribution of taxa prevalence in samples at each taxonomic level.

@@

\includegraphics[width=\textwidth,height=0.85\textheight]{figure_s4.png}\\
\textbf{Supplemental Figure 4: Top 10 important taxa at each taxonomic
level.} Summary of the 10 most important taxa for the random forest
models at each taxonomic level based on the average decrease in AUC when
the feature is permuted.

@@

\leavevmode\hypertarget{ref-lobo2008}{}%
\CSLLeftMargin{5. }
\CSLRightInline{\textbf{Lobo JM}, \textbf{Jiménez-Valverde A},
\textbf{Real R}. 2008. AUC: a misleading measure of the performance of
predictive distribution models. Global Ecology and Biogeography
\textbf{17}:145--151.
doi:\href{https://doi.org/10.1111/j.1466-8238.2007.00358.x}{10.1111/j.1466-8238.2007.00358.x}.}

\leavevmode\hypertarget{ref-callahan2016}{}%
\CSLLeftMargin{6. }
\CSLRightInline{\textbf{Callahan BJ}, \textbf{McMurdie PJ},
\textbf{Rosen MJ}, \textbf{Han AW}, \textbf{Johnson AJA}, \textbf{Holmes
SP}. 2016. DADA2: High-resolution sample inference from Illumina
amplicon data. Nature Methods \textbf{13}:581--583.
doi:\href{https://doi.org/10.1038/nmeth.3869}{10.1038/nmeth.3869}.}
